\documentclass{tufte-book}
\input{preamble.tex}
% \usetikzlibrary{external}
% \tikzexternalize

% Book metadata
\title{Truchet}
\subtitle{\texorpdfstring{$4\times4$}{4x4} patterns with four-fold rotational symmetry}
\author[]{Dan MacKinnon}
\publisher{\texorpdfstring{\url{https://github.com/dmackinnon1/Truchet-Book}}{}}

\begin{document}

% Front matter
%\frontmatter

% r.1 blank page
%\blankpage


% r.3 full title page
\maketitle


% v.4 copyright page
\newpage
\begin{fullwidth}
~\vfill
\thispagestyle{empty}
\setlength{\parindent}{0pt}
%\setlength{\parskip}{\baselineskip}
\textit{Current printing, \today}\\
\textit{Source, \url{https://github.com/dmackinnon1/Truchet-Book}}\\
\textit{Contact, \href{mailto://dmackinnon1@gmail.com}{dmackinnon1@gmail.com}}
\end{fullwidth}

\cleardoublepage

\chapter*{Introduction}

\noindent
\marginnote{\centering\input{intro_generated_files/tileList2.gtex}}A plain square has four-fold rotational symmetry: it looks the same after being rotated by $90^{\circ}$. Decorating a square can change its appearance and break this symmetry, so that the square looks different when rotated. Traditionally, Truchet squares are rendered asymmetrical by being divided by a diagonal line and coloured differently on either side of the diagonal. With this decoration, a Truchet square loses the plain square's rotational symmetry and can be rotated to four distinguishable positions. \marginnote[3\baselineskip]{\centering\input{intro_files/positions}}Patterns can be formed by tiling a surface with Truchet squares, rotating individual squares into different positions to create repeated motifs. Among the more pleasing patterns that can be made with Truchet squares are ones that express a restored symmetry that the individual squares lack.

\vspace{0.3cm}
\noindent
This booklet presents a complete listing of $4\times 4$ Truchet tiling patterns with four-fold ($90^{\circ}$) rotational symmetry ($256$ patterns).  Treating these $4\times 4$ patterns as tiles themselves allows for larger decorative patterns to be constructed from them. For example, a uniform frieze made from a single $4\times 4$ tile can actually produce interesting secondary patterns which help illustrate some interesting relationships that exist among the tile patterns.  

\vspace{0.3cm}
\noindent
\marginnote{\centering\input{intro_files/intro_example}}Each $4\times 4$
Truchet tile pattern with rotational symmetry has a core $2\times 2$ pattern in one of its quadrants that is rotated to produce the overall pattern.  In this booklet, the core pattern, or prototile, is assumed to be in the lower left. Each pattern can identified as a sequence of $4$ digits $(a,b,c,d)$, or more succinctly, $abcd$, that list the rotational positions of each square in the lower left quadrant. This sequence $abcd$ will be referred to as the \textit{signature} of the tile pattern.

\chapter{Pattern families}

\noindent
We can group $4\times 4$ Truchet tile patterns with rotational symmetry into families where tile patterns are in the same family if they are identical with the colouring removed, leaving behind just the diagonal line in each square. The squares that make up a colourless family pattern have two-fold rotational symmetry, so the signature that represents the family of a tile pattern can be found by taking the signature of the tile pattern \textit{modulo} $2$. For example, the $16$ tile patterns below are all members of the $0110$ family.  

\vspace{1.2cm}

\marginnote{\centering\input{ch1_generated_files/parent-0110.gtex}\\ \textit{The 0110 family pattern}}

{
\setlength{\tabcolsep}{3pt}
\renewcommand{\arraystretch}{2}
\input{intro_files/example_family}
{\begin{center} \textit{The 0110 pattern family}\end{center}}
}

\vspace{0.5cm}

\noindent
For a given family, there is  \textit{companion} family, the family of patterns formed by rotating each square in a member of the original family by $90^{\circ}$. In this way, the companion family represents a set of patterns that are ``as different as possible'' from the original family. On the following pages each family will be shown along with its \textit{companion} family, providing a complete listing of all $4\times 4$ Truchet tile patterns with rotational symmetry.
\newpage
\input{ch1_families}

\chapter{Family and tile pattern mappings}

\noindent
For each family, in addition to the companion family of patterns formed by rotating each square in a member of the original family by $90^{\circ}$,  there are also two \textit{skew} families, formed by taking the upper left and lower right quadrants of an original family tile pattern as a founding prototile pattern and a \textit{dual} family, formed by taking the upper right quadrant as a founding prototile pattern. A family is always different than its companion, but it can happen that the family and its skews or duals can coincide. Self-dual families, where the dual family is the same as the original, contain some interesting cases of the frieze patterns explored in the next chapter.

\section{Self-Dual families}
\input{ch1_selfDuals}

\section{Non self-dual families}
\input{ch1_notSelfDuals}

\newpage

\section{Family and tile mappings}
\marginnote{\centering\input{intro_generated_files/tileList2.gtex}} If $\mathcal{T}$ is the set of $4\times 4$ Truchet patters with four-fold rotational symmetry, and $\mathcal{F}$ is the set of $4x4$ Truchet tile families, the relationship between tile patterns and their families is a mapping $\mathcal{T} \rightarrow \mathcal{F}$. Both individual $4\times 4$ Truchet tiles patterns and their families can be described by their signatures. As previously mentioned, the signature of the family of a tile can be found by considering the tile's signature $\pmod{2}$.

\begin{align*}    
\text{family}:& \, t \mapsto F\\
\text{family}:& \, (a,b,c,d) \mapsto (a, b, c, d) \pmod{2}\\
\end{align*}

\subsection{Family mappings}
Related families ($\mathcal{F}\longrightarrow\mathcal{F}$) can be obtained from applying simple operations on the signature a pattern family $F$.

\begin{align*}        
\text{companion}: (a,b,c,d) &\mapsto (a+1, b+1, c+ 1, d+1) \pmod{2};\\
    \text{skew}+ : (a,b,c,d) &\mapsto (c+1, a+1, d+ 1, b+1) \pmod{2};\\
    \text{dual} : (a,b,c,d) &\mapsto (d, c, b, a) \pmod{2};\\
    \text{skew}- : (a,b,c,d) &\mapsto (b+1, d+1, a+ 1, c+1) \pmod{2};
\end{align*}

\subsection{Tile pattern mappings}
Mappings between related tile patterns ($\mathcal{T}\longrightarrow\mathcal{T}$) can be described in terms of simple operations on the signature of a pattern $t$.
\marginnote{\centering\input{intro_files/positions}}
\begin{align*}        
    \text{skew}+ : (a,b,c,d) &\mapsto (c+1, a+1, d+ 1, b+1) \pmod{4};\\
    \text{dual} : (a,b,c,d) &\mapsto (d+2, c+2, b+ 2, a+2) \pmod{4};\\
    \text{skew}- : (a,b,c,d) &\mapsto (b+3, d+3, a+ 3, c+3) \pmod{4};\\
    \text{opposite} : (a,b,c,d) &\mapsto (a+2, b+2, c+2, d+2) \pmod{4};
\end{align*}

\section{Some theorems}

\noindent
For a tile pattern $t$ in a family $F$ ($t\in F$), we will write $t^d$, $t^+$, $t^-$, and  $t^{op}$ for the dual, positive skew, negative skew, and opposite of $t$, respectively. Simlarly for the family $F$ we will write $F^d$, $F^+$, $F^-$, and  $F^{c}$ for the dual, positive skew, negative skew, and companion of $F$. Some simple theorems that describe the relationships among these concepts are fun to state and easy to prove.

\begin{theorem}
A tile $t$ is self-dual ($t = t^d$) if and only if its prototile has two-fold rotational symmetry.      
\end{theorem}

\begin{theorem}
A tile is self-skew ($t=t^+$ or $t=t^-$) if and only if its prototile has four-fold rotational symmetry.
\end{theorem}

\begin{theorem}
If a tile $t$ is self-skew ($t=t^+$ or $t=t^-$) then $t$ is also self-dual ($t = t^d$).
\end{theorem}

\begin{theorem}
The dual of a tile is a member of the dual of the original tile's family.
$$t\in F \implies t^d \in F^d$$  
\end{theorem}

\begin{marginfigure}
\[
\begin{tikzcd}
	{\mathcal{T}} & {\mathcal{T}} \\
	{\mathcal{F}} & {\mathcal{F}}
	\arrow["{\small\text{dual}}", from=1-1, to=1-2]
	\arrow["{\small\text{family}}"', from=1-1, to=2-1]
	\arrow["{\small\text{family}}", from=1-2, to=2-2]
	\arrow["{\small\text{dual}}"', from=2-1, to=2-2]
\end{tikzcd}
\]
\end{marginfigure}

\begin{theorem}
A self-dual tile is a member of a self dual family. 
$$t\in F, \: t=t^d \implies F = F^d$$     
\end{theorem}

\begin{theorem}
A op-dual tile is a member of a self dual family.  
$$t\in F, \: t^{op}=t^d \implies F = F^d$$     
\end{theorem}

\begin{theorem}
The companion family of a self-dual family is also self-dual.  
$$F = F^d \implies F^c = F^{cd}$$     
\end{theorem}

\begin{theorem}
A tile is self-dual if and only if its  positive and negative skew tiles are equal. 
$$t = t^d \iff t^+ = t^-$$
\end{theorem}

% \begin{theorem}
% If a tile is self-dual ($t = t^d$), then its prototile has two-fold rotational symmetry.
% \end{theorem}

% \begin{theorem}
% If a tile is self-skew ($t=t^+$ or $t=t^-$) then its prototile has four-fold rotational symmetry, $t$ is self-dual, and $t^+ = t^-$. 
% \end{theorem}

\begin{theorem}
The opposite of a tile $t$ is in the same family as $t$. 
$$t\in F \implies t^{op} \in F$$
\end{theorem}
\begin{marginfigure}[-7cm]
\[
\begin{tikzcd}
	{\mathcal{T}} & {\mathcal{T}} \\
	& {\mathcal{F}}
	\arrow["{\small\text{op}}", from=1-1, to=1-2]
	\arrow["{\small\text{family}}"', from=1-1, to=2-2]
	\arrow["{\small\text{family}}", from=1-2, to=2-2]
\end{tikzcd}
\]
\end{marginfigure}
\begin{theorem}
The dual of a tile's dual is the original tile. 
$$(t^d)^d = t$$
\end{theorem}
\begin{marginfigure}[-5cm]
\[\begin{tikzcd}
	{\mathcal{T}} & {\mathcal{T}} \\
	& {\mathcal{T}}
	\arrow["{\small\text{dual}}", from=1-1, to=1-2]
	\arrow["{{\small 1_\mathcal{T}}}"', from=1-1, to=2-2]
	\arrow["{\small\text{dual}}", from=1-2, to=2-2]
\end{tikzcd}\]
\end{marginfigure}

\begin{theorem}
The dual of a tile's opposite is the opposite of the tile's dual. 
$$(t^{op})^d = (t^d)^{op}$$
\end{theorem}
\begin{marginfigure}[-3cm]
\[
\begin{tikzcd}
	{\mathcal{T}} & {\mathcal{T}} \\
	{\mathcal{T}} & {\mathcal{T}}
	\arrow["{\small\text{dual}}", from=1-1, to=1-2]
	\arrow["{\small\text{op}}"', from=1-1, to=2-1]
	\arrow["{\small\text{op}}", from=1-2, to=2-2]
	\arrow["{\small\text{dual}}"', from=2-1, to=2-2]
\end{tikzcd}
\]
\end{marginfigure}
\begin{theorem}
There are $256$ $4\times 4$ Truchet patterns with four-fold rotational symmetry in $16$ families of $16$ patterns each. There are four self-dual families, each with four self-dual tiles. Two of the self-dual families include two self-skew tiles each.  
\end{theorem}

\newpage

\chapter{Uniform friezes}

\noindent
Each $4\times 4$ Truchet pattern can be treated like a tile and used in a larger pattern. A uniform \textit{frieze} is a horizontal strip of the same tile pattern repeated. Friezes of $4\times 4$ Truchet pattern tiles with rotational symmetry can be quite striking, and have some interesting characteristics. 

\vspace{0.5cm}
\noindent
A frieze of more than one row of a primary tile reveals a secondary tile pattern that appears as another horizontal strip of $4\times 4$ Truchet tile patterns nestled between the rows of primary tiles. Below, a frieze of 2223 tiles has a secondary pattern of 1000 tiles.
\,

\vspace{0.5cm}
\input{ch2_files/sample_frieze}

\vspace{0.5cm}
\noindent
The secondary tile in a frieze pattern is the pattern that has been referred to previously as the \textit{dual} of the original pattern. The dual of a tile pattern is the pattern formed by taking the top right quadrant of the original tile as the prototile of the new tile.

\vspace{0.5cm}
\noindent
Some tiles are self-dual, and frieze patterns formed by self-dual tiles show a much more uniform pattern, as the extra rows of tiles seemingly nestled between the rows of the original tile are made up of the same original tile. Friezes of self-dual tiles have a third \textit{tertiary} tile pattern with four-fold rotational symmetry that appears to overlap between adjacent tiles of the original tile. These tertiary tile patterns are the \textit{skew} of the original tile pattern. Some self-dual friezes are also self-skew, leading to even more uniform patterns.

\vspace{0.5cm}
\noindent
We can consider the uniform friezes formed by the dual tiles as the same frieze pattern. There are $6$ pairs of families where the original and dual are not the same, and these pairs of families yield $16$ patterns each. The $4$ remaining families contain some self-dual patterns, and some patterns that are \textit{op-dual} (the secondary tile is the opposite tile of the original), also reducing the number of patterns. These $4$ remaining families provide $10$ distinct frieze patterns each. This means that the $256$ tile patterns generate $136$ distinct friezes. 

\vspace{0.5cm}
\noindent

\noindent
\newpage

\input{ch2_friezes}

\chapter{Self-dual tiles}

\noindent
Self-dual tiles are the $4\times 4$ Truchet tile patterns whose $2\times 2$ prototile has two-fold ($180^{\circ}$) rotational symmetry. Because of the two-fold rotational symmetry of the prototile, its appearance in the third quadrant of the $4\times 4$ tile is identical to its appearance in the initial quadrant. So the dual tile that emerges when placing four of the tiles together in a larger $2\times 2$ tile array is another copy of the original tile, appearing in the center of the  larger $2\times 2$ pattern. 

\section{Self-dual tiles that are not self-skew}

If the original protitle has strict two-fold symmetry,  $2\times 2$ patterns made with the four-fold rotationally symmetrical Truchet tile also display another distinct emergent four-fold rotationally symmetrical Truchet tile, which we are calling the \textit{tertiary} tile. 

\noindent
In these patterns, it appears that there are five copies of the primary tile (four placed in a $2\times 2$ array, and another emerging in the center), along with four copies of the tertiary tile pattern.

{
\setlength{\tabcolsep}{0pt}
\renewcommand{\arraystretch}{0}
\input{ch3_selfdual_weak}
}
\section{Self-skew tiles}

\noindent
If the original protitle has four-fold symmetry, in $2\times 2$ patterns made with the tertiary tile is another copy of the original four-fold rotationally symmetrical Truchet tile. In this the pattern becomes very uniform, a $4\times4$ repeating pattern of the underlying prototile.

{
\setlength{\tabcolsep}{0pt}
\renewcommand{\arraystretch}{0}
\input{ch3_selfdual_strong}
}
\chapter{Semicircle Truchet tile patterns}

\noindent
Instead of the traditional Truchet square, any pattern that breaks the rotational symmetry of the square different placement options ($2$ or $4$) can be used. One alternative to the traditional Truchet is to cut the square by an arc, so that a quarter circle is produced instead of a right triangle. 

\,
\vspace{0.2cm}

{
\setlength{\tabcolsep}{3pt}
\renewcommand{\arraystretch}{2}
\input{ch4_generated_files/tileList.gtex}
}

\,
\vspace{0.2cm}
{
\setlength{\tabcolsep}{0pt}
\renewcommand{\arraystretch}{0}
\input{ch4_curved_selections}
}

\,
\vspace{0.2cm}

\noindent
In the version used here, the quarter circle is  white against a black background. The unequal colour distribution (favouring either white or black) makes this Truchet variation unbalanced when compared with traditional Truchet tiles. The family groupings no longer bring tile patterns together based on "forgetting" the colour, and the family resemblances among members are somewhat weaker. It is still helpful to group these tiles in the same families as the original tiles, even though the members do not resemble each other as strictly as they do in the traditional case. On the following pages, the families of $4\times4$  tiles with rotational symmetry are shown using this semicircle Truchet square variant.

\newpage

\input{ch4_families}

\chapter{Conclusions}

\noindent
The absence of rotational symmetry in the basic Truchet square provides an opportunity express symmetry in larger patterns made from these squares. Truchet tile patterns with $4\times 4$ rotational symmetry are nice to explore - aesthetically appealing, they generate a set that is small enough to completely describe yet large enough to express some interesting variety and relationships.

\,
\vspace{0.4cm}

\noindent
Looking at Truchet variants suggests that some of the relationships in traditional patterns can be preserved and others altered, opening up new areas to explore.

\backmatter
\nocite{*}
\bibliography{references}
\bibliographystyle{plainnat}
%\printindex

\end{document}
